*flippant* -- An R package for the automated analysis of dithionite scramblase

### 2016-05-03 inst/manuscript/manuscript.rmd
Homoginizing of title formatting;

#Abstract
Dithionite scramblase assays are an established means of probing the lipid 
scrambling activity of protein extracts and purified proteins. While the assay
has yielded insight into the scramblase activity in crude membrane preparations,
functional validation of candidate scramblases, stoichiometry of scramblase 
complexes and probe ATP-dependence of flippases, data analysis in its context 
has remained a task involving many manual steps. With the extension package 
*flippant* to R, a free software environment for statistical computing and 
graphics, we introduce an integrated solution for the analysis and 
publication-grade graphical presentation of dithionite scramblase assays and 
demonstrate its utility in revisiting an originally manual analysis from the 
publication record, closely reproducing the reported results.

@@

#Availablility and Implementation
*flippant* has been implemented in R [@r_core_team_r:_2015] and integrated 
into the `bioconductor` collection of tools for biomedical data analysis [@huber_orchestrating_2015]. It is thus install- and loadable as follows from 
within a functional R environment:

@@

#Input Data and Data Processing
To analyse raw data stemming from dithionite scramblase assays, *flippant* 
requires for each data point information on the path to the 
fluorometer-generated spectrum, the amount of protein used to reconstitute 
proteoliposomes, the assay volume before and after addition of 
dithionite (to calculate a volume correction factor), the amount of lipids 
present (to calculate a protein to phospholipid ratio or PPR) etc. The user 
supplies this information in a spreadsheet (see below for supporting 
functionality).

@@

#Main Features
##Template Generation for Data Input
Supporting the user in providing the input data needed, the following function 
generates a spreadsheet program-compatible tab-delimited plain text template, 
including column names, commentary, expected data type and the default used when
omitted.

@@

##Graphic Representation of Spectral Traces
In a first, graphical data analytic step of a scramblase assay-derived data 
set, fluorescent spectra or traces may be plotted in publication-appropriate
quality with acquisition time on the x- and fluorescence intensity on the y-axis
using

@@

##Scramblase Assay Analysis
Plotting protein to phospholipid ratio (PPR) against the likelihood for a given
proteoliposome to contain one or more scramblases, the PPR plot is at the heart 
of the analysis of dithionite scramblase assays, allowing for the visual 
comparison of specific scramblase activity between interrogated experimental 
series, which may, for example, be representative of protein preparations including 
mutant protein forms or crude extracts with or without depletion of a candidate
scramblase.

@@

#Case Study
*Retinitis pigmentosa* is a degenerative disease of the retina, the majority of 
cases of which is linked to mutant forms of the G-protein coupled receptor 
rhodopsin [@mendes_mechanisms_2005]. Motivated by the recent discovery that 
opsin/rhodopsin has scramblase activity [@menon_opsin_2011;@goren_constitutive_2014],
we hypothesized that a class of enigmatic rhodopsin mutations, which are known
to cause *retinitis pigmentosa* yet do not display the commonly associated 
molecular phenotypes of impaired binding of retinal, folding and/or trafficking or 
transducin activation, may in fact be impacted in their scramblase activity.
This hypothesis was disproved when tested using the dithionite scramblase assay
among other techniques. However, the same set of experiments pointed at a defect 
in rhodopsin dimerization prior to insertion into liposomes, generating the new 
hypothesis that the mutations may cause disease by interfering with the highly 
ordered quarternary structure of the molecule in the retina 
[@ploier_perturbed_submitted].

@@

#Conclusion

@@

#Funding
J.G., R.J.C. and the Proteomics Core at WCMC-Q are supported by ``Biomedical 
Research Program'' funds at Weill Cornell Medical College in Qatar, a program 
funded by Qatar Foundation. This study was made possible by NPRP grant
5-669-1-112 from the Qatar National Research Fund (a member of Qatar 
Foundation).The statements made herein are solely the responsibility of the
authors. B.P. was also supported by the Austrian Science Fund (FWF), project
J3686-B21.

#Acknowledgements

#References

>>>>

# Abstract
Dithionite scramblase assays are an established means of probing the lipid 
scrambling activity of protein extracts and purified proteins. While the assay
has yielded insight into the scramblase activity in crude membrane preparations,
functional validation of candidate scramblases, stoichiometry of scramblase 
complexes and probe ATP-dependence of flippases, data analysis in its context 
has remained a task involving many manual steps. With the extension package 
*flippant* to R, a free software environment for statistical computing and 
graphics, we introduce an integrated solution for the analysis and 
publication-grade graphical presentation of dithionite scramblase assays and 
demonstrate its utility in revisiting an originally manual analysis from the 
publication record, closely reproducing the reported results.

@@

# Availablility and Implementation
*flippant* has been implemented in R [@r_core_team_r:_2015] and integrated 
into the `bioconductor` collection of tools for biomedical data analysis [@huber_orchestrating_2015]. It is thus install- and loadable as follows from 
within a functional R environment:

@@

# Input Data and Data Processing
To analyse raw data stemming from dithionite scramblase assays, *flippant* 
requires for each data point information on the path to the 
fluorometer-generated spectrum, the amount of protein used to reconstitute 
proteoliposomes, the assay volume before and after addition of 
dithionite (to calculate a volume correction factor), the amount of lipids 
present (to calculate a protein to phospholipid ratio or PPR) etc. The user 
supplies this information in a spreadsheet (see below for supporting 
functionality).

@@

# Main Features
## Template Generation for Data Input
Supporting the user in providing the input data needed, the following function 
generates a spreadsheet program-compatible tab-delimited plain text template, 
including column names, commentary, expected data type and the default used when
omitted.

@@

## Graphic Representation of Spectral Traces
In a first, graphical data analytic step of a scramblase assay-derived data 
set, fluorescent spectra or traces may be plotted in publication-appropriate
quality with acquisition time on the x- and fluorescence intensity on the y-axis
using

@@

## Scramblase Assay Analysis
Plotting protein to phospholipid ratio (PPR) against the likelihood for a given
proteoliposome to contain one or more scramblases, the PPR plot is at the heart 
of the analysis of dithionite scramblase assays, allowing for the visual 
comparison of specific scramblase activity between interrogated experimental 
series, which may, for example, be representative of protein preparations including 
mutant protein forms or crude extracts with or without depletion of a candidate
scramblase.

@@

# Case Study
*Retinitis pigmentosa* is a degenerative disease of the retina, the majority of 
cases of which is linked to mutant forms of the G-protein coupled receptor 
rhodopsin [@mendes_mechanisms_2005]. Motivated by the recent discovery that 
opsin/rhodopsin has scramblase activity [@menon_opsin_2011;@goren_constitutive_2014],
we hypothesized that a class of enigmatic rhodopsin mutations, which are known
to cause *retinitis pigmentosa* yet do not display the commonly associated 
molecular phenotypes of impaired binding of retinal, folding and/or trafficking or 
transducin activation, may in fact be impacted in their scramblase activity.
This hypothesis was disproved when tested using the dithionite scramblase assay
among other techniques. However, the same set of experiments pointed at a defect 
in rhodopsin dimerization prior to insertion into liposomes, generating the new 
hypothesis that the mutations may cause disease by interfering with the highly 
ordered quarternary structure of the molecule in the retina 
[@ploier_perturbed_submitted].

@@

# Conclusion

@@

# Funding
J.G., R.J.C. and the Proteomics Core at WCMC-Q are supported by ``Biomedical 
Research Program'' funds at Weill Cornell Medical College in Qatar, a program 
funded by Qatar Foundation. This study was made possible by NPRP grant
5-669-1-112 from the Qatar National Research Fund (a member of Qatar 
Foundation).The statements made herein are solely the responsibility of the
authors. B.P. was also supported by the Austrian Science Fund (FWF), project
J3686-B21.

# Acknowledgements

# References
